State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources

### 2020-08-20 analysis/paper/paper.Rmd
Corrected a few spelling errors and other minor edits. Re-knit to all 3 formats

---
title: "State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources"
author:
  - Kevin E. See:
      email: Kevin.See@merck.com
      institute: [BM]
      correspondence: true
  - Ryan N. Kinzer:
      email: ryank@nezperce.org
      institute: [NezPerce]
      correspondence: false
  - Mike W. Ackerman:
      email: mike.ackerman@merck.com
      institute: [BM]
      correspondence: false
institute:
  - BM: Biomark, Inc. 705 South 8th St., Boise, Idaho, 83702, USA
  - NezPerce: Nez Perce Tribe, Department of Fisheries Resource Management, 14054 Burr Road, PO Box 1942, McCall, Idaho, 83638, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
    bookdown::pdf_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      geometry: margin=1in
      latex_engine: "xelatex"
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
      includes:
        in_header: ../templates/header_manuscript.tex
    bookdown::html_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
fontsize: 12pt
mainfont: Times New Roman
bibliography:
- references.bib
- packages.bib
csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
abstract: |
  This is the abstract
keywords: |
  escapement; Chinook; steelhead; Lower Granite Dam; state-space model; Bayesian model
highlights: |
  These are the highlights. 
---

@@

Fish escapement is the number of adults that survive juvenile and adult rearing, escape harvest and return to their natal habitat to potentially spawn (e.g. @Bue1998). For anadromous fishes, escapement is often estimated at a fixed location in a river system prior to fish reaching their spawning area. Escapement estimates facilitate effective fisheries management, and particularly, estimates of escapement for each component group (e.g., stock, population, age, origin [wild, hatchery]; @hess2014monitoring; @Steinhorst2017) provide valuable information that fisheries managers use to achieve sustainable harvest, while protecting small and vulnerable populations. Accurate escapement estimates are increasingly important for depleted populations as they inform viable salmonid metrics [@mcelhany2000viable], and facilitate assessments of population viability and extinction risk [@nwfsc2015; @williams2016viability] and climate change vulerability [@crozier2019climate].

Populations of Chinook Salmon Oncorhynchus tshawytscha and steelhead trout O. mykiss in the Snake River basin of Pacific Northwest, USA, are depleted following decades of substantial harvest and anthropogenic changes to their migration corridor (e.g., construction of hydroelectric dams on the Snake and Columbia rivers) and tributary habitats [@nehlsen1991pacific; @mcclure2003large; @nwfsc2015]. As a result, Snake River spring-summer run Chinook Salmon (hereafter Chinook Salmon) were classified as threatened in 1992 under the Endangered Species Act (ESA; Federal Registry Notice 57 FR 14653), and Snake River summer-run steelhead trout (hereafter steelhead) were listed as threatened five years later (Federal Registry Notice 62 FR 43937). Snake River Chinook Salmon and steelhead have substantial recreational, commercial, cultural and subsistence value both within the Snake River basin as well as in downstream corridors (i.e., Columbia River) and ocean fisheries. The aggregate escapement of Snake River Chinook Salmon and steelhead populations, with the exception of Tucannon River, is measured at Lower Granite Dam located in southeast Washington; the final dam on the Snake River that returning adults must pass prior to heading to tributary spawning locations. Many fisheries management and conservation actions are made based on estimates of escapement at Lower Granite Dam parsed by species and origin (**Snake Basin Harvest Plans?**). Additionally, harvest openings and closures, both upstream in Snake River fisheries and downstream in mainstem Snake and Columbia River fisheries, are predicated on escapements at Lower Granite Dam (**could cite US v OR Management Agreement and Snake Basin Harvest Plans?**).

Chinook Salmon and steelhead returning to a majority of populations in the Snake River must ascend a fish ladder on Lower Granite Dam before migrating to their natal tributary spawning locations. Fish management agencies previously used counts of fish, by species, passing an observation window located on the side of the fish ladder as a census of fish escaping past the dam. For both species, total escapement was then parsed into groups (e.g., wild, hatchery) using observed marks and genetic data from a sample of fish captured at an adult trap located on the fish ladder upstream of the observation window [@Camacho2017, @Steinhorst2017]. Window counts as a census proved beneficial as being an easy, straight-forward method that was ascertained in near real-time. Moreover, downriver fisheries management arenas have used window counts at lower-river dams as escapement estimates for the past several years, and consistency in methods is often desirable for management decision making (**citation?**).

@@

We used STADEM and three sources of data to estimate Chinook Salmon and steelhead escapement at Lower Granite Dam from 2010-2019. Data sources included 1) counts of fish migrating past the observation window located on the adult fish ladder at Lower Granite, 2) information from adults captured at a fish trap located in the fish ladder, and 3) observations of previously PIT tagged fish detected in the adult fish ladder. Following, we describe each of the data sources in more detail as they pertain to Lower Granite Dam; however, similar data could likely be obtained from other fish passage facilities.

@@

Daily counts of adult Chinook Salmon and steelhead passing an observation window located on the Lower Granite fish ladder were the first source of data. Daily counts were made and provided by the US Army Corps of Engineers, and when summed, provide an estimate of the number of fish ascending and passing (i.e., escapement) Lower Granite Dam each season. Window counts were made for each species using video monitoring and direct in-person visual monitoring during daytime hours [@Hatch1994]. Video monitoring occured during the beginning and tail ends of the adult runs (March 1 – March 31 and November 1 – December) for 10 hours per day (0600 – 1600 hours). Direct visual monitoring occured during peak run times (April 1 – October 31) for 16 hours per day (0400 – 2000 hours) [@USACE2015]. During direct visual monitoring, observers recorded each adult (≥ 30cm), by species, passing the window for 50 minutes of each hour of operation. Salmonids under 30cm in length were not identified to species. The sum of the daily 50-minute counts were then multiplied by 1.2 to account for the 10 minutes when fish were not counted. Daytime window counts were not expanded for fish that may have ascended the ladder outside of operational hours (i.e., nighttime) [@USACE2015]. Window counts were accessed through the Columbia Basin Research Data Access in Real Time (DART) website, [www.cbr.washington.edu/dart/query/adult_daily](www.cbr.washington.edu/dart/query/adult_daily), using their window count query. Counts were provided for each day the fish ladder was open to passage. Although window counts were assumed to be a census of every fish passing Lower Granite Dam, corrections were not applied for nighttime passage or re-ascending fish. Further, there was no estimation of daily or seasonal observation or sampling error.

@@

The nighttime passage rate was estimated as the count of tags that migrated through the fish ladder during non-window observation hours divided by the total number of tags passing the fish ladder, and was calculated on a weekly basis. The re-ascension rate was the count of tags observed passing the upstream most detection sites in the adult fish ladder (i.e., passing the dam) and later detected re-entering the downstream end of the fish ladder at a later time divided by the total number of tags leaving the fish ladder. Previously, we examined for differences in nighttime passage and re-ascension rates as estimated using wild fish only, versus combining hatchery and wild fish together, and found no difference. Therefore, we combine wild and hatchery PIT tagged fish observations to estimate common nighttime passage and re-ascension rates to increase sample sizes.

@@

We estimated the total number of fish crossing the dam each week, based on the window counts and the total fish passing the adult trap, while also accounting for nighttime passage, and fallback/re-ascension rates using a state-space modeling approach [@royle2008hierarchical] implemented in the STADEM package for the R statistical software [@R-Core-Team2020]. We assumed that the window counts and the estimates from the trap (fish in the trap divided by trap rate that week) were generated by processes with observation error. In the case of the trap for example, we assumed there was sampling variation and uncertainty around our estimates of the true unknown trap rate. STADEM further accounted for the proportion of fish that ascended the ladder while the counting window was closed (i.e., night), as well as for fish potentially double-counted (or more) after falling back below the dam and later re-ascending the fish ladder. Last, adult sampling data from the trap (wild, hatchery, hatchery no-clip) were used to partition the total escapement estimate by origin (Figure \@ref(fig:stadem-examp-fig). Additional model details can be found in [Appendix A](#append1). The STADEM package is available from the primary author at a [https://github.com/KevinSee/STADEM](https://github.com/KevinSee/STADEM), and requires the use of the JAGS software [@R-rjags] for Bayesian inference.

@@

Finally, we applied STADEM to data from Lower Granite Dam for Chinook Salmon and steelhead returning to the Snake River to spawn during 2010 to 2019. Window counts for both species were accessed from DART via functions within STADEM. For Chinook Salmon, a spawn year refers to adults that migrate past the dam prior to August 17 each year and spawn that late summer and fall. For steelhead, the spawn year is defined as steelhead that migrate past Lower Granite Dam starting July 1 the previous year and prior to June 30 of the given year (e.g., spawn year 2017 steelhead migrate past Lower Granite between July 1, 2016 and June 30, 2017). Data from the adult trap was made available by Idaho Department of Fish and Game, and adult PIT tag detection data within the fish passage ladder at Lower Granite Dam was accessed from the PTAGIS regional database ([https://www.ptagis.org/](https://www.ptagis.org/)).

@@

Simulation results of observed bias, sampling variation, precision, root mean squared deviation (RMSD), and coverage probabilities were qualitatively similar for hatchery (N = 70,000) and hatchery no-clip (N = 5,000) origins as observed in the wild origin (N = 25,000) comparisons. As such, only diagnostic measures of Lower Granite Dam model fits to a medium sized escapement level (e.g., wild origin escapement) are presented.

STADEM results were very similar across all scenarios (Figure \@ref(fig:rel-bias-lgd)). Estimates were unbiased, with an average relative bias of 0.2–0.3%. The CV of the estimates average 2.0–3.0%, with coverage probabilities that always exceeded 95%. We calculated RMSD as the square root of the sum of the variance of the estimate and the squared expected bias, which accounts for the size of the uncertainty in the estimator as well as its bias. The RMSD was near 500 for each scenario, representing an estimate within 2% of the true value (Table \@ref(tab:summ-tab)).

@@

We applied STADEM to data from Lower Granite Dam for Chinook Salmon and steelhead for spawn years 2010–2019. Estimates of total escapement, as well as estimates of wild, hatchery, and hatchery no-clip estimates are presented in Table \@ref(tab:lgr-tab). Coefficients of variation ranged from 3.5-7.5% for wild fish, 3.0-6.2% for hatchery fish, 4.0-8.4% for hatchery no-clip fish and 2.9-5.6% for total unique fish past Lower Granite Dam.

Weekly estimates of total escapement over Lower Granite Dam tracked the window counts and trap estimates (Figure \@ref(fig:jags-fit-fig)). STADEM point estimates were often between estimates based on window counts, and those based on the number of fish caught in the adult trap. However, when very few fish were caught in the trap, or there was more uncertainty about the trap rate that week, STADEM estimates tracked the window counts more closely, as seen in the second week of July 2014, in Figure \@ref(fig:jags-fit-fig). That year shows the utility of STADEM in dealing with missing data, as the trap was shut down for several weeks in July and August.

@@

We have presented a novel method for estimating adult salmonid escapement past a large hydroelectric facility (e.g., Lower Granite Dam) that incorporates data from window counts, fish trap, and observations of PIT tagged fish in the adult passage ladder. Our model explicitly models nighttime passage, re-ascension, and potential error in both window and trap estimates. In doing so, we demonstrated that at Lower Granite Dam, nighttime passage and re-ascension rates do not always offset each other, and assuming they do, will lead to biased escapement estimates in some years. With minor adjustments this modelling framework and the STADEM package could be applied to similar migratory species at Lower Granite Dam (e.g., fall Chinook Salmon, Pacific lamprey Lampetra tridentata), or elsewhere; provided a fish passage barrier with a counting mechanism, a trap that can be used to sample a portion of the run, and a tag observation or detection infrastructure (e.g., a PIT tag detection array or similar) exists. Our state-space model combined multiple imperfect sources of data to reduce bias in adult escapement estimates and provided more reasonable estimates of uncertainty. Accurate population or stock abundance estimates and uncertainty accounting for observation and process error can be particularly important when estimates are used or leveraged for management and conservation decisions such as population viability analyses.

@@

Although not currently set up for this, STADEM could be modified and run on a weekly basis or in near real-time to provide better in-season estimates for fisheries managers. Currently, the only roadblock to this, at Lower Granite Dam, is the identification of hatchery origin fish from phenotypically wild fish (i.e., hatchery no-clip) using genetic tissues samples (***Steele et al. 2013***) collected at the adult trap, which currently is completed post hoc after the trapping season. The inclusion of genetic information typically results in a reduction in wild escapement estimates and an associated increase in hatchery no-clip escapement. However, if in-season management decisions do not require this correction or could accept the potential bias, origin calls at the trap could be used in-season as a first approximation to escapement. Final post-hoc estimates, parsed by origin, could then be finalized at season’s end. All other data included in this model (e.g., window counts and PIT observations) are otherwise provided in almost real-time by DART. Provided the Lower Granite Dam adult fish trap database was updated and made available frequently, there are minimal obstacles for adapting the STADEM framework to provide in-season estimates of escapement.

Recently, co-managers in the Snake River basin have adopted the STADEM framework to estimate population escapement of spring/summer Chinook Salmon and steelhead past Lower Granite Dam, and returning to tributary spawning areas[@iptds2020report; @kinzer2020report]. Estimates of escapement at Lower Granite Dam, by species and origin, include known uncertainty, and are available to further parse into sex- or age-structured escapement estimates (e.g., @Camacho2017, @Schrader2013) that are important for fisheries management and productivity monitoring of wild populations. As an example, STADEM is being applied at Lower Granite to estimate the total unique wild fish migrating past the dam. Estimates of fish passing the dam are then combined with PIT tag observations at instream PIT tag detection systems throughout the Snake River basin and estimated movement or transition probabilities, simialar to @waterhouse2020bayesian, to estimate escapement to Snake River populations and locations throughout the basin [@orme2018population]. Combined, escapement estimates from STADEM and movement probability estimates provide abundance estimates to given tributaries or populations that, joined with sex and age data collected at the adult fish trap [@Hargrove2019], provides necessary information to evaluate productivity and population viability for select Snake River Chinook Salmon and steelhead groups [@iptds2020report].

Although STADEM was developed with salmonid escapement at Lower Granite Dam in mind, it could be applied to any migratory fish species at locations with similar monitoring infrastructure. Justification and infrastructure exist for applying a modified STADEM framework for fish passing Bonneville Dam, the lowest dam on the Columbia River, or Priest Rapids Dam in the upper Columbia River. Both locations currently trap a sub-sample of passing Chinook Salmon and steelhead for biological information, and use window counts as a surrogate of true escapement, however, similar problems to those observed at Lower Granite Dam, such as, unaccounted observer and sampling error, night-time passage, and re-ascension exists at the sites. Certainly, estimating an unbiased total return to the entire Columbia River basin (i.e., Bonneville Dam) and Upper Columbia River with uncertainty would benefit managers and decision making.

>>>>

---
title: "State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources"
author:
  - Kevin E. See:
      email: Kevin.See@merck.com
      institute: [BM]
      correspondence: true
  - Ryan N. Kinzer:
      email: ryank@nezperce.org
      institute: [NezPerce]
      correspondence: false
  - Mike W. Ackerman:
      email: mike.ackerman@merck.com
      institute: [BM]
      correspondence: false
institute:
  - BM: Biomark, Inc. 705 South 8th St., Boise, Idaho, 83702, USA
  - NezPerce: Nez Perce Tribe, Department of Fisheries Resource Management, 14054 Burr Road, PO Box 1942, McCall, Idaho, 83638, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
    bookdown::pdf_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      geometry: margin=1in
      latex_engine: "xelatex"
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
      includes:
        in_header: ../templates/header_manuscript.tex
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::html_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
fontsize: 12pt
mainfont: Times New Roman
bibliography:
- references.bib
- packages.bib
csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
abstract: |
  This is the abstract
keywords: |
  escapement; Chinook; steelhead; Lower Granite Dam; state-space model; Bayesian model
highlights: |
  These are the highlights. 
---

@@

Fish escapement is the number of adults that survive juvenile and adult rearing, escape harvest and return to their natal habitat to potentially spawn (e.g. @Bue1998). For anadromous fishes, escapement is often estimated at a fixed location in a river system prior to fish reaching their spawning area. Escapement estimates facilitate effective fisheries management, and particularly, estimates of escapement for each component group (e.g., stock, population, age, origin [wild, hatchery]; @hess2014monitoring; @Steinhorst2017) provide valuable information that fisheries managers use to achieve sustainable harvest, while protecting small and vulnerable populations. Accurate escapement estimates are increasingly important for depleted populations as they inform viable salmonid metrics [@mcelhany2000viable], and facilitate assessments of population viability and extinction risk [@nwfsc2015; @williams2016viability] and climate change vulnerability [@crozier2019climate].

Populations of Chinook Salmon, Oncorhynchus tshawytscha, and steelhead trout, O. mykiss, in the Snake River basin of Pacific Northwest, USA, are depleted following decades of substantial harvest and anthropogenic changes to their migration corridor (e.g., construction of hydroelectric dams on the Snake and Columbia rivers) and tributary habitats [@nehlsen1991pacific; @mcclure2003large; @nwfsc2015]. As a result, Snake River spring-summer run Chinook Salmon (hereafter Chinook Salmon) were classified as threatened in 1992 under the Endangered Species Act (ESA; Federal Registry Notice 57 FR 14653), and Snake River summer-run steelhead trout (hereafter steelhead) were listed as threatened five years later (Federal Registry Notice 62 FR 43937). Snake River Chinook Salmon and steelhead have substantial recreational, commercial, cultural and subsistence value both within the Snake River basin as well as in downstream corridors (i.e., Columbia River) and ocean fisheries. The aggregate escapement of Snake River Chinook Salmon and steelhead populations, with the exception of Tucannon River, is measured at Lower Granite Dam located in southeast Washington; the final dam on the Snake River that returning adults must pass prior to heading to tributary spawning locations. Many fisheries management and conservation actions are made based on estimates of escapement at Lower Granite Dam parsed by species and origin (**Snake Basin Harvest Plans?**). Additionally, harvest openings and closures, both upstream in Snake River fisheries and downstream in mainstem Snake and Columbia River fisheries, are predicated on escapements at Lower Granite Dam (**could cite US v OR Management Agreement and Snake Basin Harvest Plans?**).

Chinook Salmon and steelhead returning to a majority of populations in the Snake River must ascend a fish ladder on Lower Granite Dam before migrating to their natal tributary spawning locations. Fish management agencies previously used counts of fish, by species, passing an observation window located on the side of the fish ladder as a census of fish escaping past the dam. For both species, total escapement was then parsed into groups (e.g., wild, hatchery) using observed marks and genetic data from a sample of fish captured at an adult trap located on the fish ladder upstream of the observation window [@Camacho2017; @Steinhorst2017]. Window counts as a census proved beneficial as being an easy, straight-forward method that was ascertained in near real-time. Moreover, downriver fisheries management arenas have used window counts at lower-river dams as escapement estimates for the past several years, and consistency in methods is often desirable for management decision making (**citation?**).

@@

We used STADEM and three sources of data to estimate Chinook Salmon and steelhead escapement at Lower Granite Dam from 2010-2019. Data sources included 1) counts of fish migrating past the observation window located on the adult fish ladder at Lower Granite, 2) information from adults captured at a fish trap located in the fish ladder, and 3) observations of previously PIT tagged fish detected in the adult fish ladder. Below, we describe each of the data sources in more detail as they pertain to Lower Granite Dam; however, similar data could likely be obtained from other fish passage facilities.

@@

Daily counts of adult Chinook Salmon and steelhead passing an observation window located on the Lower Granite fish ladder were the first source of data. Daily counts were made and provided by the US Army Corps of Engineers, and when summed, provide an estimate of the number of fish ascending and passing (i.e., escapement) Lower Granite Dam each season. Window counts were made for each species using video monitoring and direct in-person visual monitoring during daytime hours [@Hatch1994]. Video monitoring occurred during the beginning and tail ends of the adult runs (March 1 – March 31 and November 1 – December) for 10 hours per day (0600 – 1600 hours). Direct visual monitoring occurred during peak run times (April 1 – October 31) for 16 hours per day (0400 – 2000 hours) [@USACE2015]. During direct visual monitoring, observers recorded each adult (≥ 30cm), by species, passing the window for 50 minutes of each hour of operation. Salmonids under 30cm in length were not identified to species. The sum of the daily 50-minute counts were then multiplied by 1.2 to account for the 10 minutes when fish were not counted. Daytime window counts were not expanded for fish that may have ascended the ladder outside of operational hours (i.e., nighttime) [@USACE2015]. Window counts were accessed through the Columbia Basin Research Data Access in Real Time (DART) website, [www.cbr.washington.edu/dart/query/adult_daily](www.cbr.washington.edu/dart/query/adult_daily), using their window count query. Counts were provided for each day the fish ladder was open to passage. Although window counts were assumed to be a census of every fish passing Lower Granite Dam, corrections were not applied for nighttime passage or re-ascending fish. Further, there was no estimation of daily or seasonal observation or sampling error.

@@

<!-- Don't you think it's important to explain why we need to estimate the trap rate? Because otherwise after reading the description of the adult fish trap data, I would assume the trap rate is known -->

The nighttime passage rate was estimated as the count of tags that migrated through the fish ladder during non-window observation hours divided by the total number of tags passing the fish ladder, and was calculated on a weekly basis. The re-ascension rate was the count of tags observed passing the upstream most detection sites in the adult fish ladder (i.e., passing the dam) and later detected re-entering the downstream end of the fish ladder at a later time divided by the total number of tags leaving the fish ladder. Previously, we looked for differences in nighttime passage and re-ascension rates estimated using wild fish only, versus combining hatchery and wild fish together, and found no difference. Therefore, we combine wild and hatchery PIT tagged fish observations to estimate common nighttime passage and re-ascension rates to increase sample sizes.

@@

We estimated the total number of fish crossing the dam each week, based on the window counts and the total fish passing the adult trap, while also accounting for nighttime passage, and fallback/re-ascension rates using a state-space modeling approach [@royle2008hierarchical] implemented in the STADEM package for the R statistical software [@R-Core-Team2020]. We assumed that the window counts and the estimates from the trap (fish in the trap divided by trap rate that week) were generated by processes with observation error. In the case of the trap for example, we assumed there was sampling variation and uncertainty around our estimates of the true unknown trap rate. STADEM further accounted for the proportion of fish that ascended the ladder while the counting window was closed (i.e., night), as well as for fish potentially double-counted (or more) after falling back below the dam and later re-ascending the fish ladder. Finally, adult sampling data from the trap (wild, hatchery, hatchery no-clip) were used to partition the total escapement estimate by origin (Figure \@ref(fig:stadem-examp-fig)). Additional model details can be found in [Appendix A](#append1). The STADEM package is available from the primary author at [https://github.com/KevinSee/STADEM](https://github.com/KevinSee/STADEM), and requires the use of the JAGS software [@R-rjags] for Bayesian inference.

@@

Finally, we applied STADEM to data from Lower Granite Dam for Chinook Salmon and steelhead returning to the Snake River to spawn for the spawn years 2010 to 2019. Window counts for both species were accessed from DART via functions within STADEM. For Chinook Salmon, a spawn year refers to adults that migrate past the dam prior to August 17 each year and spawn that late summer and fall. For steelhead, the spawn year is defined as steelhead that migrate past Lower Granite Dam starting July 1 the previous year and prior to June 30 of the given year (e.g., spawn year 2017 steelhead migrate past Lower Granite between July 1, 2016 and June 30, 2017). Data from the adult trap was made available by Idaho Department of Fish and Game, and adult PIT tag detection data within the fish passage ladder at Lower Granite Dam was accessed from the PTAGIS regional database ([https://www.ptagis.org/](https://www.ptagis.org/)).

@@

Simulation results of observed bias, sampling variation, precision, root mean squared deviation (RMSD), and coverage probabilities were qualitatively similar for the hatchery (N = 70,000) and hatchery no-clip (N = 5,000) origin groups as observed in the wild origin group (N = 25,000) comparisons. As such, only diagnostic measures of Lower Granite Dam model fits to a medium sized escapement level (e.g., wild origin escapement) are presented.

STADEM results were very similar across all scenarios (Figure \@ref(fig:rel-bias-lgd)). Estimates were unbiased, with an average relative bias of 0.2–0.3%. The CV of the estimates averaged 2.0–3.0%, with coverage probabilities that always exceeded 95%. We calculated RMSD as the square root of the sum of the variance of the estimate and the squared expected bias, which accounts for the size of the uncertainty in the estimator as well as its bias. The RMSD was near 500 for each scenario, representing an estimate within 2% of the true value (Table \@ref(tab:summ-tab)).

@@

We applied STADEM to data from Lower Granite Dam for Chinook Salmon and steelhead for spawn years 2010–2019. Estimates of total escapement, as well as estimates of wild, hatchery, and hatchery no-clip estimates are presented in Table \@ref(tab:lgr-tab). Coefficients of variation ranged from 2.7-24.8% for wild fish, 2.2-10.5% for hatchery fish, 3.6-11% for hatchery no-clip fish and 2.1-10.4% for total unique fish past Lower Granite Dam.

<!-- All of the high values were from 2019, and the ranges drop quite a bit if we exclude that year. Max is more like 6%. Worth trying to mention that?  -->

Weekly estimates of total escapement over Lower Granite Dam tracked the window counts and trap estimates (Figure \@ref(fig:jags-fit-fig)). STADEM point estimates were often between estimates based on window counts, and those based on the number of fish caught in the adult trap. However, when very few fish were caught in the trap, or there was more uncertainty about the trap rate that week, STADEM estimates tracked the window counts more closely, as seen in the second week of July 2014, in Figure \@ref(fig:jags-fit-fig). That year also shows the utility of STADEM in dealing with missing data, as the trap was shut down for several weeks in July and August.

@@

We have presented a novel method for estimating adult salmonid escapement past a large hydroelectric facility (e.g., Lower Granite Dam) that incorporates data from window counts, fish trap, and observations of PIT tagged fish in the adult passage ladder. Our model explicitly models nighttime passage, re-ascension, and potential error in both window and trap estimates. In doing so, we demonstrated that at Lower Granite Dam, nighttime passage and re-ascension rates do not always offset each other, and assuming they do will lead to biased escapement estimates in some years. With minor adjustments this modelling framework and the STADEM package could be applied to similar migratory species at Lower Granite Dam (e.g., fall Chinook Salmon, Pacific lamprey, Lampetra tridentata), or elsewhere; provided a fish passage barrier with a counting mechanism, a trap that can be used to sample a portion of the run, and a tag observation or detection infrastructure (e.g., a PIT tag detection array or similar) exists. Our state-space model combined multiple imperfect sources of data to reduce bias in adult escapement estimates and provided more reasonable estimates of uncertainty. Accurate population or stock abundance estimates and uncertainty accounting for observation and process error can be particularly important when estimates are used or leveraged for management and conservation decisions such as population viability analyses.

@@

Although not currently set up for this, STADEM could be modified and run on a weekly basis or in near real-time to provide better in-season estimates for fisheries managers. Currently, the only roadblock to this, at Lower Granite Dam, is the identification of hatchery origin fish from phenotypically wild fish (i.e., hatchery no-clip) using genetic tissues samples (***Steele et al. 2013***) collected at the adult trap, which currently is completed post-hoc after the trapping season. The inclusion of genetic information typically results in a reduction in wild escapement estimates and an associated increase in hatchery no-clip escapement. However, if in-season management decisions do not require this correction or could accept the potential bias, origin calls at the trap could be used in-season as a first approximation to escapement. Final post-hoc estimates, parsed by origin, could then be finalized at season’s end. All other data included in this model (e.g., window counts and PIT observations) are otherwise provided in almost real-time by DART. Provided the Lower Granite Dam adult fish trap database was updated and made available frequently, there are minimal obstacles for adapting the STADEM framework to provide in-season estimates of escapement.

Recently, co-managers in the Snake River basin have adopted the STADEM framework to estimate population escapement of spring/summer Chinook Salmon and steelhead past Lower Granite Dam, and returning to tributary spawning areas [@iptds2020report; @kinzer2020report]. Estimates of escapement at Lower Granite Dam, by species and origin, including known uncertainty, are available to further parse into sex- or age-structured escapement estimates (e.g., @Camacho2017, @Schrader2013) that are important for fisheries management and productivity monitoring of wild populations. As an example, STADEM is being applied at Lower Granite to estimate the total unique wild fish migrating past the dam. Estimates of fish passing the dam are then combined with estimated movement or transition probabilities based on PIT tag observations at instream PIT tag detection systems throughout the Snake River basin, similar to @waterhouse2020bayesian, to estimate escapement to Snake River populations and locations throughout the basin [@orme2018population]. Combined, escapement estimates from STADEM and movement probability estimates provide abundance estimates to given tributaries or populations that, joined with sex and age data collected at the adult fish trap [@Hargrove2019], provides necessary information to evaluate productivity and population viability for select Snake River Chinook Salmon and steelhead groups [@iptds2020report].

Although STADEM was developed with salmonid escapement at Lower Granite Dam in mind, it could be applied to any migratory fish species at locations with similar monitoring infrastructure. Justification and infrastructure exist for applying a modified STADEM framework for fish passing Bonneville Dam, the lowest dam on the Columbia River, or Priest Rapids Dam in the upper Columbia River. Both locations currently trap a sub-sample of passing Chinook Salmon and/or steelhead for biological information, and use window counts as a surrogate of true escapement. However, each has at least some similar problems to those observed at Lower Granite Dam, such as unaccounted observer and sampling error, night-time passage, and/or re-ascension. Certainly, estimating an unbiased total return to the entire Columbia River basin (i.e., Bonneville Dam) and Upper Columbia River with uncertainty would benefit managers and decision making.
